---
title: "Peer-graded Assignment: Course Project 2"
output: 
        html_document:
                keep_md: true
date: "2023-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Title
Assessing Impact of Severe Weather Events: Studying NOAA Storm Database, United States, 1950– 2011

## Synopsis

## Data Processing
The data for this assignment come from the course web site: [Storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

1. Download raw data
```{r Download raw data, cache = TRUE, cache.path="cache/"}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "./df.bz2")
```

2. Read raw data
```{r Read raw data, cache=TRUE, cache.path="cache/"}
library(data.table)
df <- fread("./df.bz2")
```
3. Explore the data set
```{r}
# Features of the data set
str(df)

# Number of weather events included
length(unique(df$EVTYPE))

# Statistics of weather-related fatalities
summary(df$FATALITIES)

# Statistics of weather-related injuries
summary(df$INJURIES)
```

4. Address question 1: Across the United States, which types of events (as indicated in the  
`EVTYPE` variable) are most harmful with respect to population health?

```{r To explore the most harmful weather events with respect to population health}
library(dplyr)
library(ggplot2)
library(patchwork)

# Data extraction and rank the top 5 most fatal weather events
p1 <- df %>% 
        group_by(EVTYPE) %>% 
        summarise(total_f = sum(FATALITIES)) %>% 
        arrange(desc(total_f)) %>% 
        top_n(5) %>% 
# Plot and save barplot
        ggplot() +
                geom_bar(aes(reorder(EVTYPE, total_f), total_f), 
                         stat = "identity") +
                ylim(0, 6000) +
                theme(axis.text.x = element_text(angle = 90)) +
                labs(x = "",
                     y = "Count",
                     title = "Top 5 Causes of Weather\n Fatalities in US, 1950–Nov 2011")

# Data extraction and rank the top 5 weather events causing most injuries
p2 <- df %>% 
        group_by(EVTYPE) %>% 
        summarise(total_i = sum(INJURIES)) %>% 
        arrange(desc(total_i)) %>% 
        top_n(5) %>% 
# Plot and save barplot
        ggplot() +
        geom_bar(aes(reorder(EVTYPE, total_i), total_i), 
                 stat = "identity") +
        ylim(0, 100000) +
        theme(axis.text.x = element_text(angle = 90)) +
        labs(x = "",
             y = "",
             title = "Top 5 Causes of Weather\n Injuries in US, 1950–Nov 2011")
```

4. Address question 2: Across the United States, which types of events have the greatest economic consequences?

## Results
```{r Result for question 1}
# Combine both barplots
p1 + p2
```

